Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-01-15 acm/RBF_Section.tex
New figure from Natasha + intro text from Natasha.

\begin{figure}%[htb]
\centering
\includegraphics[width=0.5\textwidth]{figures/RBFStencil}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=21$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 21 nonzero entries per row.}
\end{figure}

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{figures/RBFStencil_n32_a}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=32$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 21 nonzero entries per row.}
\end{figure}

### 2014-01-15 acm/ieee_paper.tex
New figure from Natasha + intro text from Natasha.

An other solution, and the one we pursue in this paper, is to compute
multiple SpMVs at once on matrices that have the same sparsity
patterns. Obviously not all the applications have such a property. But
important classes of applications such as graph
recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of
derivatives of Radial Basis Functions (RBF)~\cite{FLBWSC12} can use multiple
SpMVs simultaneously. In this paper in particular, we investigate the
case of the derivative of RBFs where four derivatives of four
different functions are expressed as the multiplication of four vectors
by four matrices with identical sparsity patterns leading to the
simultaneous execution of 16 SpMVs at a time. Using multiple vectors
at a time has been investigated before to densify the
computations~\cite{Im01} but this is the first time, to the 
best of our knowledge, that the
densification comes from adding both vectors and matrices.

@@

In Section~\ref{sec:rbf}, we present the computation of RBFs and how it
can be expressed as sixteen multiplications of four vectors by four sparse matrices
with identical sparsity patterns. Section~\ref{sec:model} presents an
estimation of the instruction intensity of various forms of the
computations. We show that a sixfold improvement can be
expected when computing the sixteen multiplications simultaneously to
reach a total of about 210~Gflop/s. This performance represents
approximately 10\% of the available flop/s of a Xeon Phi
coprocessor. Therefore, it is necessary to have an implementation that
performs the computation in as few instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instructions to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental results about the amount of bandwidth that can be achieved
depending on how the SpMV kernel is written. It also provides the actual performance
of the various kernels on multiple classes of matrices, some generated
for purpose of analysis, and some extracted from an RBF application. A
performance of 135~Gflop/s is achieved using RBF-generated matrices. Concluding
remarks and perspectives are provided in Section~\ref{sec:ccl}.\looseness=-1

>>>>

Another solution, pursued in this paper, is to compute multiple SpMVs at once on matrices that have the same sparsity pattern. Obviously not all the applications have such a property. However, important classes of applications such as graph recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of derivatives for solving systems of PDEs using Radial Basis Function-generated Finite Differences (RBF-FD)~\cite{FLBWSC12} can use multiple SpMVs simultaneously. Since RBF-FD is a mesh-less method, easily handling irregular geometries and local refinement with algorithmic complexity not increasing with dimension as well as high-order derivative approximations, they are rapidly gaining ground in science and engineering modeling communities \cite{Bayona13,CDNT,FoL11,FLBWSC12,SPLM}). As a result, it is of interest to develop an efficient implementation on novel computer platforms for the calculation of derivatives within the context of RBF-FD. This paper investigates the case of approximating four different derivatives, using RBF-FD, for four different functions (a common scenario in 3D fluid dynamics modeling). Since the RBF-FD differentiation matrices approximating the four different derivative operators have an identical sparsity pattern, this leads to the simultaneous execution of 16 SpMVs at a time. Using multiple vectors at a time has been investigated before to densify the computations~\cite{Im01} but, to the best of the authors’ knowledge, this is the first time that the densification comes from adding both vectors and matrices.

@@

In Section~\ref{sec:rbf}, we introduce and further motivate the RBF-FD method, giving an example of how calculating the derivatives for a common system of PDEs in fluid dynamics can be expressed as sixteen multiplications of four vectors by four sparse matrices with identical sparsity patterns.
Section~\ref{sec:model} presents an
estimation of the instruction intensity of various forms of the
computations. We show that a sixfold improvement can be
expected when computing the sixteen multiplications simultaneously to
reach a total of about 210~Gflop/s. This performance represents
approximately 10\% of the available flop/s of a Xeon Phi
coprocessor. Therefore, it is necessary to have an implementation that
performs the computation in as few instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instructions to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental results about the amount of bandwidth that can be achieved
depending on how the SpMV kernel is written. 
It also provides the actual performance of the various kernels on multiple classes of matrices, some generated for purpose of analysis, and some extracted from an application of RBF-FD. A performance of 135~Gflop/s is achieved using RBF-FD differentiation matrices. Concluding remarks and perspectives are provided in Section~\ref{sec:ccl}.\looseness=-1
